---
title: "Toronto Airbnb Analysis"
author: "Trivikram Madala"
date: "2025-05-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(janitor)
library(lubridate)
library(leaflet)
library(sf)
library(xgboost)
library(Matrix)
library(corrplot)
```

# Introduction

This report presents a comprehensive analysis of the Airbnb short-term rental (STR) market in Toronto. Using multiple data sources, including listing data, reviews, and city program summaries, we:

- Cleaned and explored the data
- Modeled price prediction using XGBoost
- Visualized spatial patterns using GeoJSON
- Evaluated model performance and feature importance
- Presented residual maps to identify over/under-valued neighborhoods

---

# Data Loading and Cleaning

```{r load-data}
listings <- read_csv("listings.csv") %>% clean_names()
reviews <- read_csv("reviews.csv") %>% clean_names()
program_summary <- read_csv("Short Term Rental Program Data (1).csv") %>% clean_names()
registrations <- read_csv("short-term-rental-registrations-data.csv") %>% clean_names()
neighbourhoods <- read_csv("neighbourhoods.csv") %>% clean_names()
toronto_sf <- st_read("toronto_neighbourhoods.geojson")
```

## Clean Listings Data

```{r clean-listings}
listings2 <- listings %>%
  filter(!is.na(price)) %>%
  mutate(price = parse_number(price)) %>%
  filter(price > 10 & price < 1500)

listings3 <- listings2 %>%
  select(id, price, neighbourhood_cleansed, latitude, longitude, room_type,
         property_type, accommodates, bedrooms, bathrooms, beds,
         number_of_reviews, number_of_reviews_ltm, availability_365,
         review_scores_rating, review_scores_cleanliness, review_scores_value) %>%
  drop_na()
```

---

# Exploratory Data Analysis

## Price Distribution

```{r price-dist}
ggplot(listings3, aes(x = price)) +
  geom_histogram(binwidth = 25, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = scales::dollar) +
  labs(title = "Distribution of Airbnb Prices", x = "Price", y = "Frequency") +
  theme_minimal(base_size = 14)
```

## Correlation Plot

```{r corr-plot}
numeric_clean <- listings3 %>%
  select(where(is.numeric)) %>%
  select(where(~ n_distinct(.) > 1))

corrplot(cor(numeric_clean), method = "color", tl.cex = 0.8, title = "Correlation Plot of Numeric Features")
```

---

# XGBoost Modeling (Raw Prices)

## Data Preparation

```{r xgb-prep}
listings_model <- listings3
X <- model.matrix(price ~ . - price - id, data = listings_model)
X_sparse <- Matrix(X, sparse = TRUE)
y <- listings_model$price
set.seed(123)
train_idx <- sample(nrow(X_sparse), 0.8 * nrow(X_sparse))
dtrain <- xgb.DMatrix(data = X_sparse[train_idx, ], label = y[train_idx])
dtest <- xgb.DMatrix(data = X_sparse[-train_idx, ], label = y[-train_idx])
```

## Model Training and Evaluation

```{r xgb-train}
params <- list(objective = "reg:squarederror", eval_metric = "rmse", eta = 0.1, max_depth = 6)
xgb_model <- xgb.train(params, dtrain, nrounds = 100, watchlist = list(train = dtrain, test = dtest),
                       early_stopping_rounds = 10, verbose = 0)

preds <- predict(xgb_model, dtest)
actuals <- y[-train_idx]
results <- tibble(actual = actuals, predicted = preds)
rmse <- sqrt(mean((results$actual - results$predicted)^2))
r2 <- 1 - sum((results$actual - results$predicted)^2) / sum((results$actual - mean(results$actual))^2)
```

## Scatter Plot

```{r scatter-plot}
ggplot(results, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.3, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Actual vs Predicted Airbnb Prices", x = "Actual Price", y = "Predicted Price") +
  theme_minimal()
```

---

# Feature Importance

```{r importance}
importance_matrix <- xgb.importance(model = xgb_model)
xgb.plot.importance(importance_matrix[1:20, ])
```

---

# Spatial Visualization: Predicted Price Map

```{r map-predicted}
listings3$predicted_price <- NA
listings3$predicted_price[-train_idx] <- preds
map_pred <- listings3 %>%
  group_by(neighbourhood_cleansed) %>%
  summarise(pred_price = mean(predicted_price, na.rm = TRUE))

toronto_pred_map <- toronto_sf %>%
  left_join(map_pred, by = c("neighbourhood" = "neighbourhood_cleansed"))

pal <- colorNumeric("YlGnBu", toronto_pred_map$pred_price, na.color = "#f0f0f0")
leaflet(toronto_pred_map) %>%
  addTiles() %>%
  addPolygons(fillColor = ~pal(pred_price), color = "#444", weight = 1,
              fillOpacity = 0.7, label = ~paste0(neighbourhood, ": $", round(pred_price))) %>%
  addLegend(pal = pal, values = ~pred_price, title = "Avg Predicted Price")
```

---

# Residual Mapping (Actual - Predicted)

```{r residual-map}
listings3$residual <- listings3$price - listings3$predicted_price
resid_map <- listings3 %>%
  group_by(neighbourhood_cleansed) %>%
  summarise(avg_resid = mean(residual, na.rm = TRUE))

toronto_resid_map <- toronto_sf %>%
  left_join(resid_map, by = c("neighbourhood" = "neighbourhood_cleansed"))

pal2 <- colorBin("RdBu", domain = toronto_resid_map$avg_resid, bins = 7, reverse = TRUE)
leaflet(toronto_resid_map) %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(fillColor = ~pal2(avg_resid), color = "white", weight = 1,
              fillOpacity = 0.7, label = ~paste0(neighbourhood, ": ", round(avg_resid))) %>%
  addLegend(pal = pal2, values = ~avg_resid, title = "Avg Residual (Actual - Predicted)")
```

---

# Conclusion

- Our model achieved an RMSE of ~88 and R² of ~0.59.
- Top predictors include: accommodates, bathrooms, bedrooms.
- Central neighborhoods tend to be overvalued compared to model predictions.

This project illustrates the potential for using open urban data to guide pricing, policy enforcement, and equity analysis in Toronto’s dynamic rental market.
